Convergence rate of MAP estimates \\ for the exponential family

### 2021-09-30 main.tex
Add details to procimal point proof

\section{Stochastic Bregman Proximal Point}
\subsection{Correspondence with MAP}
For a dataset of independent samples $\cD_n = \{X_1,\dots,X_n\}$, the posterior can be seen iteratively
\begin{align}
	p(\natp |\cD_n)
	=p(\natp|X_n,\cD_{n-1})
	\propto p(X_n|\natp,\cD_{n-1}) p(\natp|\cD_{n-1})
	=p(X_n|\natp)p(\natp |\cD_{n-1}) \; .
\end{align}
For the MAP of the exponential family with a Bregman conjugate prior, the posterior takes the form
\begin{align}
	- \log p(\natp | \cD_n) 
	&= \sum_{i=1}^n -\log p(X_i | \natp) - \log p(\natp) + \cst\\
	&= \sum_i - \langle T(X_i) ; \natp \rangle  + n \logpart(\natp) + n_0 \bregman(\natp ; \natp_0) +\cst \\
	&= (n+n_0)\logpart(\natp)  - \langle n_0 \mu_0 + \sum_i T(X_i) ; \natp \rangle + \cst \\
	&= (n+n_0) \bregman(\natp; \hat \natp_n^\text{MAP}) + \cst \; ,
\end{align}
so we immediately get the recursion
\begin{align}
    \hat \natp_n^\text{MAP}
    &= \argmin_\natp - \log p(\natp | \cD_n) \\
    &= \argmin_\natp - \log p(X_n|\natp) - \log p(\natp | \cD_{n-1}) \\
    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
    \label{eq:SBPP}
\end{align} 
Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0}$.

@@

\paragraph{Bregman Proximal Point}
This algorithm was first published in \citet{eckstein1993nonlinear} with an asymptotic convergence property. \citet{chen1993convergence} then provided a finite sample convergence rate for deterministic Bregman proximal point. Their analysis relies on the three points lemma (Appendix~\ref{app:3points})
\begin{align}
	 \forall \natp, \lr_t f(\natp) + \cB(\natp ; \natp_t) \geq \lr_t f(\natp_{t+1}) + \cB( \natp; \natp_{t+1}) +\cB(\natp_{t+1} ; \natp_{t}) \; .
\end{align}
Applying it with $\natp=\natp_t$, we prove that function values are decreasing
\begin{align}
	f(\natp_{t+1})  \leq f(\natp_{t})  - \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \; .
\end{align}
Then applying it with $\natp=\natp^*$ (assuming such a minimizer exists), we obtain a telescopic sum
\begin{align}
	\lr_t(f(\natp_{t+1}) - f(\natp^*)) +  \cB(\natp_{t+1} ; \natp_{t}) 
	&\leq  \cB( \natp^*; \natp_{t})  - \cB( \natp^*; \natp_{t+1}) \\
	\implies \sum_{t=0}^{T-1} \lr_t (f(\natp_{t+1}) - f(\natp^*)) 
	+ \sum_{t=0}^{T-1} \cB(\natp_{t+1} ; \natp_{t}) 
	&\leq \cB( \natp^*; \natp_{0})  - \cB( \natp^*; \natp_{T})  
	\leq \cB( \natp^*; \natp_{0})  \; ,
\end{align}
which proves on the way that the total path length $ \sum_{t=0}^{t-1} \cB(\natp_{t+1} ; \natp_{t}) $ is bounded.
Using the decreasing property of $f(\natp_t)$, we get
\begin{align}
	f(\natp_{T}) - f(\natp^*) \leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t} \; .
\end{align}
This upper bound converges to $0$ whenever the sum of step-sizes converges to $0$. 
This rate can be arbitrarily fast because the oracle is strong. Indeed, taking an inifinite step-size solves the problem with one iteration, but the update is (theoretically) harder to solve for larger step-sizes.

\paragraph{Stochastic Setting.}
While \citet{eckstein1998approximate} shows some asymptotic convergence of the Bregman proximal point algorithm when each iteration is only solved approximately, we are not aware of published analysis when iterations are stochastic.
The deterministic proof does not transpose immediately to the stochastic setting.
First off, we do not have a decreasing sequence anymore
\begin{align}
	f(\natp_{t+1}, X_t)  \leq f(\natp_{t}, X_t)  
	- \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
	\implies \expect{f(\natp_{t+1}, X_t)} \leq f(\natp_{t}) 
	- \inv{\lr_t}\expect{\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}))}  \; .
\end{align}
Indeed the inequality bears on $\expect{f(\natp_{t+1}, X_t)}$, which features a correlation between $\natp_{t+1}$ and $X_t$, rather than $f(\natp_{t+1})$.
Similarly, applying the three points lemma with $\natp^*$, taking the tower expectation, and telescoping the sum, we get
\begin{align}
	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_t) - f(\natp^*)}
	\leq \cB( \natp^*; \natp_{0})  \; .
\end{align}
To pursue further, we need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
The most straightforward such assumption is 
\begin{align}
\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t})} \leq \lr_t \sigma^2 \; ,
\label{eq:proximal_variance}
\end{align}
e.g. we relate the stochastic function value on the next iterate to the function value.
Then, taking the iterate with the least expected loss, we get the rate
\begin{align}
	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
\end{align}
which is very analogous to standard rates for SGD.
It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
With the sums  $\sum \lr_t$ and $\sum \lr_t^2$, we would need a $O(\inv{\sqrt{t}})$ step-size to ensure a $O(\inv{\sqrt{t}})$ convergence rate, but then the iterate becomes a weighted average of the sufficient statistics, giving more weight to the end of the tail. This may be suited for online learning, but not for our stationary setting.

>>>>

\section{Stochastic Bregman Proximal Point}
\subsection{Correspondence with MAP}
For a dataset of independent samples $\cD_n = \{X_1,\dots,X_n\}$, the posterior can be seen iteratively
\begin{align}
	p(\natp |\cD_n)
	=p(\natp|X_n,\cD_{n-1})
	\propto p(X_n|\natp,\cD_{n-1}) p(\natp|\cD_{n-1})
	=p(X_n|\natp)p(\natp |\cD_{n-1}) \; .
\end{align}
For the MAP of the exponential family with a Bregman conjugate prior, the posterior takes the form
\begin{align}
	- \log p(\natp | \cD_n) 
	&= \sum_{i=1}^n -\log p(X_i | \natp) - \log p(\natp) + \cst\\
	&= \sum_i - \langle T(X_i) ; \natp \rangle  + n \logpart(\natp) + n_0 \bregman(\natp ; \natp_0) +\cst \\
	&= (n+n_0)\logpart(\natp)  - \langle n_0 \mu_0 + \sum_i T(X_i) ; \natp \rangle + \cst \\
	&= (n+n_0) \bregman(\natp; \hat \natp_n^\text{MAP}) + \cst \; ,
\end{align}
so we immediately get the recursion
\begin{align}
    \hat \natp_n^\text{MAP}
    &= \argmin_\natp - \log p(\natp | \cD_n) \\
    &= \argmin_\natp - \log p(X_n|\natp) - \log p(\natp | \cD_{n-1}) \\
    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
    \label{eq:SBPP}
\end{align} 
Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0}$. As a reminder this algorithm is defined with the update
\begin{align}
	 \hat \natp_n
    &= \argmin_\natp \lr_n f(\natp; X_{n}) + \bregman(\natp; \hat \natp_{n-1}) \\
    \iff \nabla \logpart(\natp_{n-1}) &= \nabla\logpart(\natp_n) + \lr_t \nabla f(\natp_n ; X_t) \; .
\end{align}
In words, if it converges to $\natp^*$, it can be seen as a backward mirror ascent starting from $\natp^*$.

@@

\subsubsection{Deterministic Bregman Proximal Point}
Deterministic Bregman proximal point. was first published in \citet{eckstein1993nonlinear} with an asymptotic convergence property, soon completed by a finite sample convergence rate in \citet{chen1993convergence}. 
This convergence rate relies on the three points lemma (Appendix~\ref{app:3points})
\begin{align}
	 \forall \natp, \lr_t f(\natp) + \cB(\natp ; \natp_t) \geq \lr_t f(\natp_{t+1}) + \cB( \natp; \natp_{t+1}) +\cB(\natp_{t+1} ; \natp_{t}) \; .
\end{align}

\paragraph{Monotonicity of Function Values.}
Applying it with $\natp=\natp_t$, we prove that function values are decreasing
\begin{align}
	f(\natp_{t+1})  \leq f(\natp_{t})  - \inv{\lr_t} \underbrace{\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) }_{=: \cS(\natp_{t} ; \natp_{t+1})}
\end{align}
where we introduced $\cS$ the symmetrized Bregman divergence
\begin{align}
	\cS(x; y) = \cB(x ; y) + \cB(y ; x) = \langle\nabla A(x) - \nabla A(y) ; x-y \rangle \; .
\end{align}

\paragraph{Descent Lemma and Telescopic Sum.}
Applying the three point lemma with $\natp=\natp^*$ (assuming such a minimizer exists), we obtain a telescopic sum
\begin{align}
	\lr_t(f(\natp_{t+1}) - f(\natp^*)) +  \cB(\natp_{t+1} ; \natp_{t}) 
	&\leq  \cB( \natp^*; \natp_{t})  - \cB( \natp^*; \natp_{t+1}) \\
	\implies \sum_{t=0}^{T-1} \lr_t (f(\natp_{t+1}) - f(\natp^*)) 
	+ \sum_{t=0}^{T-1} \cB(\natp_{t+1} ; \natp_{t}) 
	&\leq \cB( \natp^*; \natp_{0})  - \cB( \natp^*; \natp_{T})  
	\leq \cB( \natp^*; \natp_{0})  \; ,
\end{align}
which proves on the way that the total path length $ \sum_{t=0}^{t-1} \cB(\natp_{t+1} ; \natp_{t}) $ is bounded.
Using the monotonicity of $f(\natp_t)$, we get
\begin{align}
	f(\natp_{T}) - f(\natp^*) \leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t} \; .
\end{align}
This upper bound converges to $0$ whenever the sum of step-sizes converges to $0$. 
This rate can be arbitrarily fast because the oracle is strong. Indeed, taking an inifinite step-size solves the problem with one iteration, but the update is (theoretically) harder to solve for larger step-sizes.

\subsubsection{Stochastic Bregman Proximal Point.}
While \citet{eckstein1998approximate} shows some asymptotic convergence of the Bregman proximal point algorithm when each iteration is only solved approximately, we are not aware of published analysis when iterations are stochastic.

\paragraph{Variance Assumption.}
The deterministic proof does not transpose immediately to the stochastic setting.
We need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
As we will see soon, the most straightforward such assumption is 
\begin{align}
\boxed{\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t+1})} \leq \lr_t \sigma^2 \; , }
\label{eq:proximal_variance}
\end{align}
e.g. we relate the stochastic function value on the next iterate to the function value.
Asymptotically, this assumption is equivalent to the one of~\citet{hanzely2018fastest}. 
To see this, linearize $f(\natp_{t+1})$ and $f(\natp_{t+1} ; X_{t+1}) $ in $\natp_t$
\begin{align}
	\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t+1})} 
	&\approx \expect{f(\natp_{t}) - f(\natp_{t} ; X_{t+1})} + \expect{\langle \nabla f(\natp_t) - \nabla f(\natp_t ; X_{t+1}) ; \natp_{t+1} - \natp_t \rangle } \\
	&= 0 + \expect{\langle \nabla f(\natp_t) - \nabla f(\natp_t ; X_{t+1}) ; \natp_{t+1} \rangle } \\
	& = \Cov(- \nabla f(\natp_t ; X_{t+1})  ; \natp_{t+1} ) \; .
\end{align}

\paragraph{Monotonicity of Expected Function Values.}
Applying the tree point lemma with $\natp=\natp_t$ gives
\begin{align}
	f(\natp_{t+1}, X_{t+1})  
	&\leq f(\natp_{t}, X_{t+1})  - \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
	&= f(\natp_{t}, X_{t+1})  - \inv{\lr_t} \cS(\natp_{t+1} ; \natp_{t}) \\
	\implies \expect{f(\natp_{t+1}, X_{t+1})} 
	&\leq f(\natp_{t}) - \inv{\lr_t}\expect{\cS(\natp_{t+1} ; \natp_{t})}  \; .
	\label{eq:SPP_monotone}
\end{align}
This inequality bears on $\expect{f(\natp_{t+1}, X_{t+1})}$, which features a correlation between $\natp_{t+1}$ and $X_{t+1}$, and is different from $f(\natp_{t+1})$.
Combining~\eqref{eq:SPP_monotone} with the variance assumption gives
\begin{align}
	\expect{f(\natp_{t+1})} 
	&\leq \expect{f(\natp_{t+1} ; X_{t+1})}  + \lr_t \sigma^2 \\
	&\leq f(\natp_t) +  \lr_t \sigma^2 - \inv{\lr_t}\expect{\cS(\natp_{t+1} ; \natp_{t})}
\end{align}
from which we see that monotonicity is guaranteed if
\begin{align}
	\sigma^2 \leq \inv{\lr_t^2}\expect{\cS(\natp_{t+1} ; \natp_{t})} \; .
\end{align}
This is quite annoying, as our variance assumption requires $\sigma^2$ to be large, and this inequality requires $	\sigma^2 $ to be small -- the step we take should be greater than the variance.

\paragraph{Descent Lemma and Telescopic Sum.}
Similarly, applying the three points lemma with $\natp^*$, and taking te expectation with respect to $X_{t+1}$ yields
\begin{align}
	\lr_t(\E[f(\natp_{t+1},X_{t+1})] - f(\natp^*)) +  \E[\cB(\natp_{t+1} ; \natp_{t})] 
	&\leq  \cB( \natp^*; \natp_{t})  - \E[\cB( \natp^*; \natp_{t+1})] \; .
\end{align}

taking the tower expectation, and telescoping the sum, we get
\begin{align}
	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_{t+1}) - f(\natp^*)}
	\leq \cB( \natp^*; \natp_{0})  \; .
\end{align}
Then, taking the iterate with the least expected loss, we get the rate
\begin{align}
	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
\end{align}
which is very analogous to standard rates for SGD.
It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
With the sums  $\sum \lr_t$ and $\sum \lr_t^2$, we would need a $O(\inv{\sqrt{t}})$ step-size to ensure a $O(\inv{\sqrt{t}})$ convergence rate, but then the iterate becomes a weighted average of the sufficient statistics, giving more weight to the end of the tail. This may be suited for online learning, but not for our stationary setting.
